Efficient Hybrid MPI+OpenMP Programming

### 2024-07-14 HPC/paper.tex
Added introduction

%todo: look into title
\title{Parallelising Hybrid MPI Programms efficiently}
\author{Dreßler Justus}
\orcid{0009-0005-0657-5854}
\affiliation{
    \institution{Friedrich Schiller University}
    \department{Faculty of Mathematics and Computer Science}
    \streetaddress{Fürstengraben 1, Main University Building (Universitätshauptgebäude)}
    \postcode{07743}
    \city{Jena}
    \country{Germany}
}
\email{justus.dressler@uni-jena.de}

@@

% Historisch MPI-Everywhere
% heute MPI+OpenMP (warum)
% Was sind die schwierigkeiten mit MPI+OpenMP
% Paper versucht Lösungen zusammenzufassen
Message Passing Interface is the de facto standard for scaling High Performance Computing (HPC) Workloads across nodes.
Historically applications relied on the memory wasteful model of MPI-Everywhere to achieve good scalability across nodes.
In the last decades however a rapid increase in core counts per CPU neccessitated a switch to MPI+OpenMP to better use newer CPU's ressources.
While this did work, it brought fourth new challenges in designing software for HPC workloads.
The main pain point is the much harder to achieve parallel communication in MPI+OpenMP compared to MPI-Everywhere.
The newest MPI version (4.0) introduced new ways to expose communication parallelism in MPI software to ease this pain point.
So this paper tries to give an overview over the current ways to expose parallel communication to build scalable MPI+OpenMP software.
This includes a brief theoretical look at each method as well as sample code so developers may apply these methods to their own software.

% this feels like making a buzzword list

% Problemstellung: Wie expose ich logisch Parallele Nachrichten an eine MPI Biblithek (MPICH, OpenMPI) sodass diese es auf parallele Kommunikationswege mappen kann?
% behandelt nur CPU, keine Acceleratoren
% reminder: Einleiten was MPI ist, was MPI Bibliotheken sind, populäre nennen (MPICH, OpenMP, IntelMPI?), sagen was OpenMP ist und wofür es genutzt wird, Warum man es zusammen tun sollte,

\section{Motivation}
The historically preferred method of MPI-Everywhere made logical parallelism easy to express in software.
%todo: cite or rework
It did however have massive problems with memory management.
Classical MPI-Everywhere programs spawn a single process per core available.
MPI+OpenMP programs normally spawn a single process per node or NUMA domain \cite{zambreLessonsLearned2022}.

%todo: cite MPI 3 Standard
A MPI communicator is a construct that defines which processes may talk with each other.
Each MPI process is required to keep a list of all participating MPI processes in the default communicator \verb|MPI_COMM_WORLD|.
Assuming an $n$ core processor $n$ copies of the list are lying in each nodes memory in the MPI-Everywhere model.
An equivalent MPI+OpenMP program shares its list between its cores and also spawns $n$ times less processes, so \verb|MPI_COMM_WORLD| is $n$ times smaller and $n$ times less often on each processor.
For current generation processors exceeding 100 cores like Intel® Xeon® 6 \cite{intelXeon6} or 4th Generation AMD EPYC™ \cite{amd4thGenEpyc} this equates to a reduction of \verb|MPI_COMM_WORLD|'s memory consumption of over $n^2 > 10.000$ times.

@@

MPI operations like \verb|MPI_Send| take a \verb|<Rank,Tag,Communicator>| triple as input.
The tag is sent to differentiate messages between the same processes on a single communicator.
Normally the MPI standard requires a non-overtaking order for all messages with the same \verb|<Rank,Communicator>| \cite{mpi40}.
So the default settings do not enable a program to use the additionally sent tags for parallelisation.
If the program however guarantees MPI, that it will not use the \verb|MPI_ANY_TAG| (a parameter matching all tags), MPI libraries can send different messages between the same ranks but differing tags in arbitrary order.

>>>>

%todo: look into title
\title{Efficient Hybrid MPI+OpenMP Programming}
\author{Dreßler Justus}
\orcid{0009-0005-0657-5854}
\affiliation{
    \institution{Friedrich Schiller University}
    \department{Faculty of Mathematics and Computer Science}
    \streetaddress{Fürstengraben 1, Main University Building (Universitätshauptgebäude)}
    \postcode{07743}
    \city{Jena}
    \country{Germany}
}
\email{justus.dressler@uni-jena.de}

@@

% Historisch MPI-Everywhere
% heute MPI+OpenMP (warum)
% Was sind die schwierigkeiten mit MPI+OpenMP
% Paper versucht Lösungen zusammenzufassen
Message Passing Interface is the de facto standard for scaling High Performance Computing (HPC) Workloads across multiple compute nodes.
This paper will speak of MPI as the standard and refer to MPI implementations like MPICH and OpenMPI as "MPI libraries" going forward.
These MPI libaries implement the MPI standard while trying to achieve maximum performance while adhearing to it.
To do so some libraries include extra library specific settings as hints, which the library user can set to increase possible performance.
Sometimes these libraries will also include entirely new ways to parallelise communication like MPIX\_Stream \cite{Zhou2022}, but this paper will focus mainly on the MPI standard itself.

Traditionally applications relied on the memory-hungry MPI-Everywhere model \cite{zambreLessonsLearned2022}.
In the last decades however a rapid increase in core counts per CPU relative to other on-node resources neccessitated a switch to a model of MPI+X,
where MPI is used to parallise between different nodes and X is used to parallelise on each node.
Some popular choices for this second X part are CPU multithreading libraries like OpenMP or POSIX Threads, GPU Acceleration with CUDA or using MPI with Remote Memory Access calls.
This paper will focus on MPI+OpenMP, though many aspects of this paper also apply for any other CPU multithreading library.

OpenMP is a popular library for generating shared-memory multithreaded programs.
It works under the fork join model, where a main thread does all the serial work and additional threads get spawned for multithreading when needed.
It is easy to add to existing serial code, mostly by adding comments called pragmas in front of loops.
Its ease of use also makes it popular for developing HPC software.
MPI+OpenMP programs tend to perform worse than their MPI-Everywhere counterparts \cite{zambreLessonsLearned2022}\cite{zambreLogicalParallel2021}.
The main problem being extra unneeded thread synchronization overheads \cite{zambreLessonsLearned2022} and less exposure of parallelism for MPI libraries to use.

This paper gives an overview of ways to expose parallelism in communication to MPI libaries to achieve maximum performance.
For this it will discuss mainly point-to-point communication. It compares traditional methods like generating multiple MPI communicators with the newer features of MPI 4.0, which introduced new hints to make tags a viable method of exposing parallelisation and added new partitioned operations which can gather contributions of multiple threads in parallel.

\section{Motivation}
The historically preferred method of MPI-Everywhere made logical parallelism easy to express in software.
It does however use up more memory and other on chip ressources relative to MPI+OpenMP.
Classical MPI-Everywhere programs spawn a single process per core available.
MPI+OpenMP programs normally spawn a single process per node or NUMA domain \cite{zambreLessonsLearned2022}.

MPI point-to-point communication runs on so called communicators.
A MPI communicator defines which processes may talk with each other.
Each MPI process is required to keep a list of all participating MPI processes in the default communicator \verb|MPI_COMM_WORLD| \cite{mpi40}.
Assuming an $n$ core processor $n$ copies of the list are lying in each nodes memory in the MPI-Everywhere model.
An equivalent MPI+OpenMP program shares its list between its cores and also spawns $n$ times less processes, so \verb|MPI_COMM_WORLD| is $n$ times smaller and $n$ times less often on each processor.
For current generation processors exceeding 100 cores like Intel® Xeon® 6 \cite{intelXeon6} or 4th Generation AMD EPYC™ \cite{amd4thGenEpyc} this equates to a reduction of \verb|MPI_COMM_WORLD|'s memory consumption of over $n^2 > 10.000$ times.

@@

MPI operations like \verb|MPI_Send| take a \verb|<Rank,Tag,Communicator>| triple as input.
The tag is included to differentiate messages between the same processes on a single communicator.
Normally the MPI standard requires a non-overtaking order for all messages with the same \verb|<Rank,Communicator>| \cite{mpi40}.
So the default settings do not enable a program to use the additionally sent tags for parallelisation.
If the program however guarantees MPI, that it will not use the \verb|MPI_ANY_TAG| (a parameter matching all tags), MPI libraries can send different messages between the same ranks but differing tags in arbitrary order.
